fit_crcvm

### 2025-03-21 R/application/local/CRCVM/fit_crcvm.Rmd
update code for application to narwhal data

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
library(here)
library(xtable)
library(sf)
library(gganimate)
library(doParallel)
library(foreach)
library(ggpubr)

@@

  # Fit baseline
  baseline <- SDE$new(formulas = formulas,
                              data = data,type = "CRCVM_SSM",
                              response = response,par0 = c(2,1,a,b,D0,D1,sigma_D,sigma_theta),
                              fixpar=fixpar,
                              other_data=
                                list("H"=H,"interpolated_distance"=interpolated_dist,
                                     "interpolated_angle"=interpolation_data$BoundaryAngle))
  baseline$update_lambda(c(1,1))
  baseline$fit()

@@

#' Make latex table to show the estimates and confidence intervals
#' @param sde fitted sde
#' @param include_sigma_obs boolean
#' @param n_post number of draws from the posterior distribution used to give estimates
#' and confidence intervals
#' @return NULL (just print latex table)
make_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,n_post=100) {

  est=get_baseline_estimates(sde,include_sigma_obs,n_post)
  mean_par=est$estimates
  sd_par=est$std.error
  quant_par=est$quantiles
  parameter_names <- names(mean_par)
  estimates <- mean_par
  conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
          quant_par[[param]][2])
    })

@@

#' Make plot to compare BoundaryDistance in simulation and observation
#' @param simulated_data list of data frames simulated from an SDE model, each containing a column "BoundaryDistance".
#' @param observed_data real GPS data (data frame).
#' @return ggplot of boundary distances distributions
make_distance_density_plot <- function(simulated_data, observed_data) {

@@

@@

  # Plot density distribution
  plot_combined <- ggplot(combined_data, aes(x = BoundaryDistance, color = Source, fill = Source)) +
    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
    xlab("Distance to Shore") +
    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
    theme_minimal() +
    theme(legend.title = element_blank())

  return(plot_combined)
}

#' Make plot to compare BoundaryAngle in simulation and observation
#' @param simulated_data data simulated from a sde model containing a column "BoundaryAngle".
#' @param observed real GPS data
#' @return ggplot of boundary angle distributions
make_angle_density_plot <- function(simulated_data, observed_data) {

  simulated_combined$BoundaryAngle<-abs(simulated_combined$BoundaryAngle)
  observed_combined$BoundaryAngle<-abs(observed_combined$BoundaryAngle)
  combined_data <- bind_rows(simulated_combined, observed_combined)

  # Plot density distribution
  plot_combined <- ggplot(combined_data, aes(x = BoundaryAngle, color = Source, fill = Source)) +
    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
    xlab(expression(Theta)) +
    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
    theme_minimal() +
    theme(legend.title = element_blank())

  return(plot_combined)

@@

#' Make plot to compare speed in simulation and observation
#' @param simulated_data data simulated from a sde model containing a column "speed".
#' @param observed real GPS data
#' @return ggplot of boundary angle distributions
make_speed_density_plot <- function(simulated_data, observed_data,response) {

  # Plot density distribution
  plot_combined <- ggplot(combined_data, aes(x = speed, color = Source, fill = Source)) +
    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
    geom_density(data = observed_combined, alpha = 0.2) +  # Single density for observed data
    xlab("Speed") +
    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
    theme_minimal() +
    theme(legend.title = element_blank())

  return(plot_combined)
}
```

@@

```{r}
D0=quantile(dataBE12_prep200$BoundaryDistance,0.8)
D1=quantile(dataBE12_prep200$BoundaryDistance,0.5)
a=0.3*D0

baseline_crcvm200<- fit_baseline(dataBE12_prep200,interpolation_bas_200,
                    b=0.5,a=a,sigma_D=1,D0=D0,D1=D1,sigma_theta=pi/6,
                    sigma_obs=0.05)

@@

```{r}
make_final_baseline_tab(baseline_crcvm200,include_sigma_obs=FALSE,n_post=10000)
```

# Check baseline models

@@

```{r subsample simulations}
#subsample to 5 minutes and add measurement noise
all_sim_sub<-lapply(all_sim,function(data) {

  data_sub<-data[seq(1,nrow(data),by=60),]
  noise<-rmvn(nrow(data_sub),rep(0,2),diag(rep(0.05^2,2)))

  data_sub[,c("x","y")]=data_sub[,c("x","y")]+noise
  return (data_sub)
})

@@

```{r animate observed vs simulated trajectories}
dataBE12_prep200$type <- "Observed"
all_sim_sub[[1]]$type <- "Simulated"

combined_data <- rbind(dataBE12_prep200[,c("BoundaryDistance","BoundaryAngle","ID","time","x","y","type")], all_sim_sub[[1]])

@@

plot_distance<-make_distance_density_plot(all_sim_sub,dataBE12_prep200) 
plot_angle<-make_angle_density_plot(all_sim_sub,dataBE12_prep200)
plot_speed<-make_speed_density_plot(all_sim_sub,dataBE12_prep200)

final_densities_plots<-ggarrange(plotlist=list(plot_distance,plot_angle),
                     ncol=2, nrow=1, common.legend = TRUE, legend="bottom", vjust=2)
ggsave("BoundaryDistance_density.pdf",path="baseline_crcvm_200",plot=plot_distance)
ggsave("BoundaryAngle_density.pdf",path="baseline_crcvm_200",plot=plot_angle)
ggsave("speed_density.pdf",path="baseline_crcvm_200",plot=plot_speed)
```

@@

colnames(dataAE_prep200)[colnames(dataAE_prep100)=="DistanceShore200"]<-"BoundaryDistance"
colnames(dataAE_prep200)[colnames(dataAE_prep100)=="theta200"]<-"BoundaryAngle"

#remove points on land
dataAE_prep200<-dataAE_prep200[dataAE_prep200$BoundaryDistance>0,]

>>>>

library(smoothSDE)
library(ggplot2)
library(dplyr)
library(plotly)
library(htmlwidgets)
library(mgcv)
library(here)
library(xtable)
library(sf)
library(gganimate)
library(doParallel)
library(foreach)
library(ggpubr)
library(fitdistrplus)
```

@@

  # Fit baseline
  baseline <- SDE$new(formulas = formulas,
                              data = data,type = "CRCVM_SSM",
                              response = response,par0 = c(1.2,4.5,a,b,D0,D1,sigma_D,sigma_theta),
                              fixpar=fixpar,
                              other_data=
                                list("H"=H,"interpolated_distance"=interpolated_dist,
                                     "interpolated_angle"=interpolation_data$BoundaryAngle))
  baseline$update_lambda(c(1,1))
  baseline$fit()

@@

#' Make latex table to show the estimates and confidence intervals
#' @param sde fitted sde
#' @param include_sigma_obs boolean
#' @param n_post number of draws from the posterior distribution used to give estimates
#' and confidence intervals
#' @return NULL (just print latex table)
make_final_baseline_tab<-function(sde,include_sigma_obs=FALSE,n_post=100,resp=FALSE) {

  est=get_baseline_estimates(sde,include_sigma_obs,n_post,resp)
  mean_par=est$estimates
  sd_par=est$std.error
  quant_par=est$quantiles
  parameter_names <- names(mean_par)
  estimates <- mean_par
  conf_intervals <- sapply(parameter_names, function(param) {
  sprintf("$[%.2f; %.2f]$", quant_par[[param]][1], 
          quant_par[[param]][2])
    })

@@

#' Make plot to compare distributions in simulation and observation
#' @param simulated_data list of data frames simulated from an SDE model, each containing columns "BoundaryDistance", "BoundaryAngle" and "speed"
#' @param observed_data real GPS data (data frame).
#' @return ggplot of boundary distances distributions
make_density_plots <- function(simulated_data, observed_data) {

@@

  # drop trajectories that reached land

@@

  # Plot density distribution
  plot_distance <- ggplot(combined_data, aes(x = BoundaryDistance, color = Source, fill = Source)) +
    geom_density(data = simulated_combined, aes(group = SimulationID), alpha = 0.1) + 
    geom_density(data = observed_combined, alpha = 0.2) +
    xlab("Distance to Shore") +
    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
    theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 20),
    axis.title.y = element_blank(),
    axis.title.x = element_blank())
    theme(legend.title = element_blank())

    plot_angle <- ggplot(combined_data, aes(x = abs(BoundaryAngle), color = Source, 
                                            fill = Source)) +
    geom_density(data = simulated_combined, aes(group = SimulationID), 
                 alpha = 0.1) + 
    geom_density(data = observed_combined, alpha = 0.2) + 
    xlab(expression(Theta)) +
    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
    theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 20),
    axis.title.y = element_blank(),
    axis.title.x = element_blank())
    theme(legend.title = element_blank())

    plot_speed <- ggplot(combined_data, aes(x = speed, color = Source, 
                                            fill = Source)) +
    geom_density(data = simulated_combined, aes(group = SimulationID), 
                 alpha = 0.1) + 
    geom_density(data = observed_combined, alpha = 0.2) + 
    xlab("Speed") +
    scale_color_manual(values = c("Simulated" = "blue", "Observed" = "red")) + 
    scale_fill_manual(values = c("Simulated" = "blue", "Observed" = "red")) +
  theme(
    axis.title = element_text(size = 16),
    axis.text = element_text(size = 16),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 20),
    axis.title.y = element_blank(),
    axis.title.x = element_blank())
    theme(legend.title = element_blank())

  return(list("distance"=plot_distance,"speed"=plot_speed,"angle"=plot_angle))
}

@@

@@

```{r}
D0=quantile(dataBE12_prep200$BoundaryDistance,0.9)
D1=quantile(dataBE12_prep200$BoundaryDistance,0.75)
a=0.22*D0

baseline_crcvm200<- fit_baseline(dataBE12_prep200,interpolation_bas_200,
                    b=2,a=a,sigma_D=1,D0=D0,D1=D1,sigma_theta=pi/6,
                    sigma_obs=0.05)

@@

```{r}
make_final_baseline_tab(baseline_crcvm200,include_sigma_obs=FALSE,n_post=100)
```

# Check baseline model

@@

## Simulate time steps

```{r reproduce realistic time steps,warning=FALSE,message=FALSE}

dtimes<-unlist(lapply(unique(dataBE12_prep200$ID),function(id) {
    sub_data<-dataBE12_prep200[dataBE12_prep200$ID==id,]
    dt=diff(sub_data$time)
    dt
}))

fit_lognorm <- fitdist(dtimes * 60, "lnorm")
fit_gamma <- fitdist(dtimes * 60, "gamma")
gofstat(list(fit_lognorm,fit_gamma))

```{r}
# Extract estimated parameters
meanlog <- fit_lognorm$estimate["meanlog"]
sdlog <- fit_lognorm$estimate["sdlog"]

# Generate new time steps
dtimes_sim <- rlnorm(nrow(dataBE12_prep200)-6, meanlog, sdlog)

times_sim<-lapply(unique(dataBE12_prep200$ID),function(ID) {
  sub_time<- dataBE12_prep200[dataBE12_prep200$ID==ID,"time"]
  sub_n<-length(sub_time)
  min_time=min(sub_time)
  times<-rep(min_time,sub_n)+c(0,cumsum(dtimes_sim[1:(sub_n-1)]/60))
  return (times)})
names(times_sim)<-unique(dataBE12_prep200$ID)
```

## Subsample and interpolate simulated trajectory

```{r}
all_sim_sub=list()

for (i in seq_along(all_sim)) {
  sim_data<-all_sim[[i]]
  interp_sim_data<-dataBE12_prep200[,c("time","ID","x","y")]
  interp_sim_data$time<-unlist(times_sim)

  for (id in unique(sim_data$ID)) {
    sub_sim_data<-sim_data[sim_data$ID==id,]
    sim_x_interp <- approx(sub_sim_data$time, sub_sim_data$x, xout = times_sim[[id]])$y
    sim_y_interp <- approx(sub_sim_data$time, sub_sim_data$y, xout = times_sim[[id]])$y
    sub_ind=interp_sim_data$ID==id
    interp_sim_data[sub_ind,"x"]<-sim_x_interp
    interp_sim_data[sub_ind,"y"]<-sim_y_interp
  }

  all_sim_sub[[i]]<-interp_sim_data
}
```

```{r subsample simulations}
#subsample to 5 minutes and add measurement noise
all_sim_sub<-lapply(all_sim_sub,function(data) {

  data<-data[!(is.na(data$x)),]
  noise<-rmvn(nrow(data),rep(0,2),diag(rep(0.02^2,2)))

  data[,c("x","y")]=data[,c("x","y")]+noise
  return (data)
})

@@

```{r animate observed vs simulated trajectories}
dataBE12_prep200$type <- "Observed"
all_sim_sub[[2]]$type <- "Simulated"

combined_data <- rbind(dataBE12_prep200[,c("ID","time","x","y","type")], all_sim_sub[[2]])

@@

plots<-make_density_plots(all_sim_sub,dataBE12_prep200)

final_densities_plots<-ggarrange(plotlist=list(plots$distance,plots$angle),
                     ncol=2, nrow=1, common.legend = TRUE, legend="bottom", vjust=2)
ggsave("BoundaryDistance_density.pdf",path="baseline_crcvm_200",plot=plots$distance)
ggsave("BoundaryAngle_density.pdf",path="baseline_crcvm_200",plot=plots$angle)
ggsave("speed_density.pdf",path="baseline_crcvm_200",plot=plots$speed)
```

@@

colnames(dataAE_prep200)[colnames(dataAE_prep200)=="DistanceShore200"]<-"BoundaryDistance"
colnames(dataAE_prep200)[colnames(dataAE_prep200)=="theta200"]<-"BoundaryAngle"

dataAE_prep200$speed<-compute_speed(dataAE_prep200)
#remove points on land
dataAE_prep200<-dataAE_prep200[dataAE_prep200$BoundaryDistance>0,]
